---
title: "Reproducible Research, project 2 : a study of storms impact in the US"
author: "Mathieu Besancon"
date: "Sunday, September 14, 2014"
output: html_document
---

##Synopsis

The report was created for the Reproducible Research course on the Coursera plateform (see presentation [here](https://www.coursera.org/course/repdata). The analysis was performed using R and RStudio. More information [on the R project](http://www.R-project.org/). 

The purpose of the analysis is to use the data available on storms in the USA to draw some conclusions on the storms having the biggest impacts on people's health and on the economy. 
  
The data were collected by the U.S. National Oceanic and Atmospheric Administration (NOAA).    
  
Pareto analyses were performed to detect the key events types for each problem. 11 key event types are highlighted for the issue of human fatalities, while only 5 event types are relevant considering the human fatalities and injuries.  
  
## Data processing
The raw data are loaded into `storm_data`, we can take a look at the size and content of the dataset.
```{r,cache=TRUE}
storm_data<-read.csv("repdata-data-StormData.csv")
dim(storm_data)
names(storm_data)
```
Since our dataset is huge, we will reduce it to some subsets for each specific need of the analysis. We create two subsets, `harm_data` for the consequences on human health and `eco_data` for the economic impact.
```{r}
use_att<-c("EVTYPE","FATALITIES","INJURIES")
harm_data<-storm_data[use_att]
eco_att<-c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
eco_data<-storm_data[eco_att]
```

###Loading the libraries
The principal library we have to load is ggplot2, which is used to create more complex plots.
The attribute "EVTYPE" describes the type of event. Let's look at the
```{r,echo=FALSE}
library(ggplot2)
```
The multiplot function will also be used for plotting several graphs at a time. The code source is available on [Cookbook for R](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/).
```{r,echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

### Studying the impact of storms on people's health  

For the impact on health, we consider two metrics as relevant, the amount of fatalities and the amount of fatalities and injuries per event (which we compute in a new attribute, `Harm`. We create two aggregations of the dataset,`agg_harm` for fatalities and `agg2` for the overall harms.

```{r}
harm_data$Harm<-harm_data$FATALITIES + harm_data$INJURIES
attach(harm_data)
agg_harm<-aggregate(FATALITIES~EVTYPE,FUN=sum)
agg2<-aggregate(Harm~EVTYPE,FUN=sum)
detach(harm_data)
```
We want to work only on events which affected fatalities and harms respectively, so we sort and clean the aggregated data, keeping only the interesting event types.
```{r}
# For fatalities
agg_harm<-agg_harm[order(-agg_harm$FATALITIES),]
agg_harm<-agg_harm[which(agg_harm$FATALITIES>0),]
total_fatal<-sum(agg_harm$FATALITIES)

# For all harms
agg2<-agg2[order(-agg2$Harm),]
agg2<-agg2[which(agg2$Harm>0),]
total_harm<-sum(agg2$Harm)
```
The cumulated percentage of the metric of interest can then be computed. We also find the observation index for which the cumulated effects reach 80%, `index_80` and `index_h` respectively.
```{r}
#For fatalities
agg_harm$Cumulated<-NA
agg_harm$Cumulated[1]<-agg_harm$FATALITIES[1]/total_fatal
agg_harm$Index<-1
index_80<-NULL
for (i in 2:(as.numeric((length(agg_harm$Cumulated))))){
        agg_harm[i,3]<-agg_harm[i-1,3]+(agg_harm[i,2]/total_fatal)
        agg_harm[i,4]<-i
}
for (i in 2:length(agg_harm$Cumulated)){
        if (agg_harm[i,3]>0.8){
                index_80<-(i)
                break
        }
}
#For all harms
agg2$Cumulated<-NA
agg2$Cumulated[1]<-agg2$Harm[1]/total_harm
agg2$Index<-1
index_h<-NULL
for (i in 2:(as.numeric((length(agg2$Cumulated))))){
        agg2[i,3]<-agg2[i-1,3]+(agg2[i,2]/total_harm)
        agg2[i,4]<-i
}
for (i in 2:length(agg2$Cumulated)){
        if (agg2[i,3]>0.8){
                index_h<-(i)
                break
        }
}
```
We represent four plots, two per metric, one pareto curve (`par1` and `par3`) and a bar plot of the main sources for each metric (`par2` and `par4`). Note : we define here the critical causes as the first causes generating 80% of the overall effect (fatalities or harms). 

```{r}
par1<-ggplot(data=agg_harm,aes(Index,Cumulated))
par1<-par1+geom_line(ylab="Cumulated percentage of fatalities",col='red',size=1.1)
par1<-par1+labs(y="Cumulated fraction of fatalities",title="Pareto analysis for fatalities")
par1<-par1+xlim(c(0,NA))+ylim(c(0,1))
par1<-par1+geom_abline(intercept=0,slope=(1/length(agg_harm$Cumulated)))
par1<-par1+geom_abline(intercept=0.8,slope=0,col='orange',size=0.5)
par1<-par1+geom_vline(xintercept=as.numeric(index_80),col='orange',size=0.5)

#Creating the subset containing just the critical event types
agg_crit<-agg_harm[1:index_80,]
par2<-qplot(weight=FATALITIES,reorder(EVTYPE,-FATALITIES),data=agg_crit,geom="bar",ylab="Fatalities",fill=Index,title="Main causes of fatalities")

par3<-ggplot(data=agg2,aes(Index,Cumulated))
par3<-par3+geom_line(ylab="Cumulated percentage of harms",col='red',size=1.1)
par3<-par3+labs(y="Cumulated fraction of harms",title="Pareto analysis for harms")
par3<-par3+xlim(c(0,NA))+ylim(c(0,1))
par3<-par3+geom_abline(intercept=0,slope=(1/length(agg2$Cumulated)))
par3<-par3+geom_abline(intercept=0.8,slope=0,col='orange',size=0.5)
par3<-par3+geom_vline(xintercept=as.numeric(index_h),col='orange',size=0.5)

agg_crit2<-agg2[1:index_h,]
par4<-qplot(weight=Harm,reorder(EVTYPE,-Harm),data=agg_crit2,geom="bar",ylab="Harms",fill=Index,title="Main causes of harms")
```
We first plot the graphs representing just the fatalities.
```{r}
multiplot(par1,par2,ncol=2)
```
The second pair of plot is taking into account the fatalities and injuries.
```{r}
multiplot(par2,par3,ncol=2)
```

### Studying the economic impact of storms

Our working data are still quite big, we delete the observations for which the cost was zero.
```{r}
eco_data<-eco_data[which(eco_data$PROPDMG>0|eco_data$CROPDMG>0),]
```

We aggregate the data with the cost exponent for crops and property damages, so that overall costs can easily be computed.

```{r}
agg_eco<-aggregate(cbind(PROPDMG,CROPDMG)~EVTYPE+PROPDMGEXP+CROPDMGEXP,data=eco_data,FUN=sum)
agg_eco$overall<-0
for (i in 1:length(agg_eco[,1])){
        if(toupper(agg_eco$PROPDMGEXP[i])=="M"){
                agg_eco$PROPDMG[i]<-agg_eco$PROPDMG[i]*10^6
        }else if (toupper(agg_eco$PROPDMGEXP[i])=="K"){
                agg_eco$PROPDMG[i]<-agg_eco$PROPDMG[i]*10^3
        }else if (toupper(agg_eco$PROPDMGEXP[i])=="B"){
                agg_eco$PROPDMG[i]<-agg_eco$PROPDMG[i]*10^9
        }
        if(toupper(agg_eco$CROPDMGEXP[i])=="M"){
                agg_eco$CROPDMG[i]<-agg_eco$CROPDMG[i]*10^6
        }else if (toupper(agg_eco$CROPDMGEXP[i])=="K"){
                agg_eco$CROPDMG[i]<-agg_eco$CROPDMG[i]*10^3
        }else if (toupper(agg_eco$CROPDMGEXP[i])=="B"){
                agg_eco$CROPDMG[i]<-agg_eco$CROPDMG[i]*10^9
        }
        agg_eco$overall[i]<-agg_eco$CROPDMG[i]+agg_eco$PROPDMG[i]
}
agg_eco<-agg_eco[order(-agg_eco$overall),]


agg_eco$Cumulated<-0
agg_eco$Index<-1
index_cost<-NULL
total_cost<-sum(agg_eco$overall,na.rm=TRUE)
agg_eco$Cumulated[1]<- agg_eco$overall[1]/total_cost
for (i in 2:length(agg_eco$overall)){
        agg_eco$Cumulated[i]<-(agg_eco$overall[i]/total_cost + agg_eco$Cumulated[i-1])
        agg_eco$Index[i]<-i
}

for (i in 2:length(agg_eco$Cumulated)){
        if (agg_eco$Cumulated[i]>0.8){
                index_eco<-(i)
                break
        }
}
```
The Pareto curve for the overall costs versus `EVTYPE` can be computed.
```{r}
par5<-ggplot(data=agg_eco,aes(Index,Cumulated))
par5<-par5+geom_line(ylab="Cumulated percentage of costs",col='red',size=1.1)
par5<-par5+labs(y="Fraction of overall costs",title="Pareto analysis for costs")
par5<-par5+xlim(c(0,NA))+ylim(c(0,1))
par5<-par5+geom_abline(intercept=0,slope=(1/length(agg_eco$Cumulated)))
par5<-par5+geom_abline(intercept=0.8,slope=0,col='orange',size=0.5)
par5<-par5+geom_vline(xintercept=as.numeric(index_eco),col='orange',size=0.5)

agg_crit3<-agg_eco[1:index_eco,]
par6<-qplot(weight=overall,reorder(EVTYPE,-overall),data=agg_eco,geom="bar",ylab="Costs ($)",fill=Index,title="Main sources of cost among storms")

multiplot(par5,par6,ncol=2)
```













